 

### 2023-05-02 markdown/Thesis.tex
normalization started to ellaborate

in partial fulfilment of the requirement \\for the degree of MSc \\in Bioinformatics

@@

\hypertarget{cell-isolation}{%
\paragraph{Cell Isolation}\label{cell-isolation}}

\hypertarget{lysis}{%
\paragraph{Lysis}\label{lysis}}

\hypertarget{digestion}{%
\paragraph{Digestion}\label{digestion}}

\hypertarget{ionization}{%
\paragraph{Ionization}\label{ionization}}

\hypertarget{coupled-mass-spectrometry-msms}{%
\paragraph{Coupled mass-spectrometry
(MS/MS)}\label{coupled-mass-spectrometry-msms}}

\hypertarget{labeling-techniques}{%
\paragraph{Labeling techniques}\label{labeling-techniques}}

@@

\hypertarget{ms-spectrum}{%
\subsubsection{MS-Spectrum}\label{ms-spectrum}}

@@

\hypertarget{protein-assembly}{%
\subparagraph{Protein assembly}\label{protein-assembly}}

After these calculation the identified peptides can be aggregated
according to it`s respective protein and quantified. The mentioned
metrics indicating the performance of the peptide search can be used in
downstream analysis. A so called razor peptide indicates the group with
the highest number of identified peptides. Quantification is enabled by
taking only unique peptides into contrast. Posterior error
probabilities, which refers to the chance that the found peptide is a
random event, are multiplied and only distinct sequences with the
highst-scoring are used.

\hypertarget{data-processing}{%
\paragraph{Data processing}\label{data-processing}}

Further analysis can be done with R and respective packages such as
bioconductor. Since there is no state of the art established, analysis
varies upon experimental design.

\hypertarget{reading-the-data}{%
\subparagraph{Reading the data}\label{reading-the-data}}

After processing MaxQuant creates a directory containing all results as
.txt file. The evidence.txt file include all peptide sequence matches
(PSM) with their respective proteins and statistical parameters.

@@

@@

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\vspace{0.5cm}

Analysis of the data was done with MaxQuant \citep{Cox2008}

>>>>

in partial fullfilment of the requirement \\for the degree of MSc \\in Bioinformatics

@@

\hypertarget{cell-isolation}{%
\subsubsection{Cell Isolation}\label{cell-isolation}}

\hypertarget{lysis}{%
\subsubsection{Lysis}\label{lysis}}

\hypertarget{digestion}{%
\subsubsection{Digestion}\label{digestion}}

\hypertarget{labeling-techniques}{%
\subsubsection{Labeling techniques}\label{labeling-techniques}}

@@

\hypertarget{mass-spectrometry-1}{%
\paragraph{Mass Spectrometry}\label{mass-spectrometry-1}}

\hypertarget{ionization}{%
\subparagraph{Ionization}\label{ionization}}

\hypertarget{ms.1}{%
\subparagraph{MS.1}\label{ms.1}}

\hypertarget{coupled-mass-spectrometry-msms-ms.2}{%
\subparagraph{Coupled mass-spectrometry (MS/MS) \&
MS.2}\label{coupled-mass-spectrometry-msms-ms.2}}

Acquisition of the data was done with MaxQuant \citep{Cox2008} software
package.

\hypertarget{ms-spectrum}{%
\paragraph{MS-Spectrum}\label{ms-spectrum}}

@@

\hypertarget{protein-assembly}{%
\paragraph{Protein assembly}\label{protein-assembly}}

After these calculation the identified peptides can be aggregated
according to it`s respective protein and quantified. The mentioned
metrics indicating the performance of the peptide search can be used in
downstream analysis. A so called razor peptide indicates the group with
the highest number of identified peptides. Quantification is enabled by
taking only unique peptides into contrast. Posterior error
probabilities, which refers to the chance that the found peptide is a
random event, are multiplied and only distinct sequences with the
highest-scoring are used.

\hypertarget{data-processing}{%
\subsubsection{Data processing}\label{data-processing}}

Further analysis is done with R and respective packages such as
bioconductor. Since there is no state of the art established, analysis
varies upon experimental design. The workflow of the analysis will be
processed in a so called pipeline streamlining the data through steps
where individual results can be observed in visualizations and indivdual
calculations will be adapted according to user demands and experimental
properties.

\includegraphics{Thesis_files/figure-latex/data_processing_pipeline_flowchart_vertical-1.pdf}

\hypertarget{reading-the-data}{%
\paragraph{Reading the data}\label{reading-the-data}}

After processing MaxQuant creates a directory containing all results as
.txt file. The evidence.txt file include all peptide to spectrum matches
(PSM) with their respective proteins and statistical parameters.

@@

\hypertarget{peptide-spectrum-match-psm-aggregation-to-peptides}{%
\paragraph{Peptide spectrum match (PSM) aggregation to
peptides}\label{peptide-spectrum-match-psm-aggregation-to-peptides}}

In data science, aggregation refers to a row-wise operation that merges
data based on a particular column using a specific function. In the
context of processing from peptides to spectrum matches, the desired
column is the peptide sequence. To account for different distributions
across multiple assays, the median of the channel is used as the
function to aggregate multiple matches into one.

\hypertarget{join-assays-when-observing-multiple-comparable-batches-at-once}{%
\paragraph{Join assays when observing multiple comparable batches at
once}\label{join-assays-when-observing-multiple-comparable-batches-at-once}}

Sample size is often a limiting factor in hypothesis testing. A strict
quality control and the fact that TMT reagents are only available up to
18-plex can reduce the number of observed samples below the critical
threshold, leading to an early end of analyses. To overcome this
limitation, the provided software is capable of processing multiple runs
simultaneously, allowing for testing of multiple batches and increasing
the number of samples that can be included in the analysis.

\hypertarget{calculate-reporter-ion-intensity-ri-and-filter-according-to-median-ri}{%
\paragraph{Calculate reporter ion intensity (RI) and filter according to
median
RI}\label{calculate-reporter-ion-intensity-ri-and-filter-according-to-median-ri}}

Columns which do not meet the desired intensity can be filtered by a
threshold set on the RI. The median RI can also be used to check if an
entire channel has a lower detection level. This can be due two reasons.
One is the expression level of the given proteins in a cell. Meaning,
that the expression of the observed cell type is simply lower than the
other type. Another one could be a spillage of TMT detection in other
channels due incorrect or missing correction of the TMT isotopes.

\hypertarget{calculate-and-filter-according-to-median-coefficient-of-variation-cv-per-cellchannel}{%
\paragraph{Calculate and filter according to median coefficient of
variation (CV) per
cell/channel}\label{calculate-and-filter-according-to-median-coefficient-of-variation-cv-per-cellchannel}}

Depending having a bulk sample or single-cell sample choosing a minimum
of observed peptides and a cutoff value for the CV, changes the level of
confidence in the peptide data. The coefficient of variation of a
peptide is considered as the ratio of the standard deviation to the mean
and describes the relationship of the observed peptide signal over
multiple proteins (=razor proteins). Peptides having a high coefficient
of variation over many razor proteins are considered as noise and need
to be filtered out before statistical analysis.

\hypertarget{remove-peptides-with-high-missing-rate}{%
\paragraph{Remove peptides with high missing
rate}\label{remove-peptides-with-high-missing-rate}}

Although missing value imputation can be performed during the analysis
of multiple batches, peptides with missing detections across channels
can be problematic for quantification. The proteomic composition of a
biological sample is similar between replicates and even across groups.
However, the threshold of missingness (described as a fraction of the
row) can be set in the user interface and adjusted to enable different
experimental designs.

\hypertarget{aggregation-of-peptides-to-proteins}{%
\paragraph{Aggregation of peptides to
proteins}\label{aggregation-of-peptides-to-proteins}}

Similar to the already explained previous aggregation step the peptides
will be further processed into their respective proteins after the
quality control on the peptide level is performed. Finally an expression
matrix for every protein and their respective channel column is
returned. Each channel refers to a sample and reflects contains the
intensities of each protein found in the biological specimen.

\hypertarget{transformation-of-protein-expression-data}{%
\paragraph{Transformation of protein expression
data}\label{transformation-of-protein-expression-data}}

Data transformation applies a function to each value of a matrix or
array, so that: \[
y_i = f(x_i)
\] Depending on the distribution of the values in the observed
expression set, different transformations can be applied to fulfill
assumptions for statistical testing. In the shiny application, various
procedures are implemented and can be further expanded upon request from
the user. The macrophage analysis done by Specht et al.~mentioned in
their SCoPE2 publication \citep{Specht2021} uses the logarithm to the
base 2 to spread a compacted distribution and remove skewness in the
dataset. This transformation was used as a reference when comparing
methods. However, using the logarithm to the base 10 may be an easily
interpretable way of defining expression data and will also be
facilitated by other transformation methods such as the boxcox method,
which is also implemented in the application. When observing a wide
distribution of small and large values of the expression set in a
histogram, a square root transformation can help increase the
variability of smaller values and decrease the variability of larger
values. Practically impossible, but in data-driven science, occasionally
negative expression values may be present, and they can be converted
into positive values by taking each one to the power of two. Depending
on the distribution, the user has to decide which transformation to
consider and verify that statistical assumptions are met after
application with visualizations such as histograms, qqplots, and
MA-plots. For statistically inexperienced users, the boxcox
transformation can help with this decision. Developed in 1964 by Box and
Cox \citep{Sakia1992}, this method applies a linear model against 1 on
the data to determine the statistical parameter lambda by the maximum
likelihood method. The log-likelihood method takes the 95\% confidence
interval for the parameter lambda and the final lambda is chosen as the
value with the highest log-likelihood value.

\begin{tabular}{l|r}
\hline
& lambda & Transformation\$\\
\hline
-2 & $\frac{1}{x^2}$
\hline
-1 & $\frac{1}{x}$
\hline
-0.5 & $\frac{1}{\sqrt{x}}$
\hline
0 & $\log{x}$
\hline
0.5 & $\sqrt{x}$
\hline
1 & $x$
\hline
2 & $x^2$
\end{tabular}

Depending on the size of lambda, a certain function will be
automatically applied to each value of the expression set in order to
introduce normality. In terms of usability, the boxcox method is the
most efficient transformation method, taking the decision of which
calculation to apply away from the user.

\hypertarget{normalization}{%
\paragraph{Normalization}\label{normalization}}

The term normalization is ambiguous in data-science and will be
explained briefly in this chapter. Starting with converting the data to
the Z-distribution (Gaussian result), where 0 reflects the mean and 1
refers to the standard deviation, until vectorization where every
feature (=protein) can be explained as a vector which points in a
specific direction in the unit sphere, normalization can have different
meanings. Normalization is considered as part of the scaling procedure
and ensures that every feature contributes equally to our statistical
model and hinders large values from biasing the model in a particular
direction. Although this could also be harmful for the modeling
procedure, since the impact of important features on the dataset will
also be reduced. Depending on the algorithms later used in the data
processing, the method of choice for normalization can change the
results significantly. The K-means clustering algorithm for instance
relies on the distances between distinct data points by minimizing
variance of the squared euclidean distances. This method will be applied
in the missing value imputation which will be the next step applied to
our data. Hence we want to make sure that the get the maximum accuracy
for our imputed values. The first option to choose in the user interface
is the column-wise median and row-wise mean normalization. This method
was chosen by Specht et al in the SCoPE2 publication 2021
/citep\{Specht2021\} and worked as a reference during the development
process to benchmark other methods. When dividing each value of an
expression set by the median of the particular column, the procedure is
considered as a column-wise median normalization. Hence the
mean-normalization works the same way on the column. In the analysis
pipeline both procedures were applied on the dataset. However when the
logarithm as the transformation method was chosen in advance the
pipeline automatically switches from a division to a subtraction, since:
\(\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v\)

@@

### 2023-05-02 markdown/methods.Rmd
normalization started to ellaborate

```{r data_processing_pipeline_flowchart, echo=FALSE, fig.height=17, fig.width=13}
library(Gmisc, quietly = TRUE)
library(glue)
library(htmlTable)
library(grid)
library(magrittr)

@@

```{r data_processing_pipeline_flowchart2, echo=FALSE, fig.height=9, fig.width=32}
library(Gmisc, quietly = TRUE)
library(glue)
library(htmlTable)
library(grid)
library(magrittr)

@@

#### Transformation of protein expression data
Data transformation applies a function to each value of a matrix or array, so that: 
$$
y_i = f(x_i)
$$
Depending on the distribution of the values in the observed expression set, different transformations can be applied to fulfill assumptions for statistical testing. In the shiny application, various procedures are implemented and can be further expanded upon request from the user. The macrophage analysis done by Specht et al. mentioned in their SCoPE2 publication \citep{Specht2021} uses the logarithm to the base 2 to spread a compacted distribution and remove skewness in the dataset. This transformation was used as a reference when comparing methods. However, using the logarithm to the base 10 may be an easily interpretable way of defining expression data and will also be facilitated by other transformation methods such as the boxcox method, which is also implemented in the application.
When observing a wide distribution of small and large values of the expression set in a histogram, a square root transformation can help increase the variability of smaller values and decrease the variability of larger values.
Practically impossible, but in data-driven science, occasionally negative expression values may be present, and they can be converted into positive values by taking each one to the power of two.
Depending on the distribution, the user has to decide which transformation to consider and verify that statistical assumptions are met after application with visualizations such as histograms, qqplots, and MA-plots.
For statistically inexperienced users, the boxcox transformation can help with this decision. Developed in 1964 by Box and Cox \citep{Sakia1992}, this method applies a recursion on the data to determine the statistical parameter lambda.

@@

>>>>

```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=17, fig.width=13}
library(Gmisc, quietly = TRUE)
library(glue)
library(htmlTable)
library(grid)
library(magrittr)

@@

```{r data_processing_pipeline_flowchart_horizontal, eval=FALSE, fig.height=9, fig.width=32, include=FALSE}
library(Gmisc, quietly = TRUE)
library(glue)
library(htmlTable)
library(grid)
library(magrittr)

@@

#### Transformation of protein expression data
Data transformation applies a function to each value of a matrix or array, so that: 
$$
y_i = f(x_i)
$$
Depending on the distribution of the values in the observed expression set, different transformations can be applied to fulfill assumptions for statistical testing. In the shiny application, various procedures are implemented and can be further expanded upon request from the user. The macrophage analysis done by Specht et al. mentioned in their SCoPE2 publication \citep{Specht2021} uses the logarithm to the base 2 to spread a compacted distribution and remove skewness in the dataset. This transformation was used as a reference when comparing methods. However, using the logarithm to the base 10 may be an easily interpretable way of defining expression data and will also be facilitated by other transformation methods such as the boxcox method, which is also implemented in the application.
When observing a wide distribution of small and large values of the expression set in a histogram, a square root transformation can help increase the variability of smaller values and decrease the variability of larger values.
Practically impossible, but in data-driven science, occasionally negative expression values may be present, and they can be converted into positive values by taking each one to the power of two.
Depending on the distribution, the user has to decide which transformation to consider and verify that statistical assumptions are met after application with visualizations such as histograms, qqplots, and MA-plots.
For statistically inexperienced users, the boxcox transformation can help with this decision. Developed in 1964 by Box and Cox \citep{Sakia1992}, this method applies a linear model against 1 on the data to determine the statistical parameter lambda by the maximum likelihood method. The log-likelihood method takes the 95% confidence interval for the parameter lambda and the final lambda is chosen as the value with the highest log-likelihood value. 
```{r lambda_table, echo=FALSE}
# 
# table <- tibble(lambda=c(-2,-1,-0.5,0,0.5,1,2), Transformation=c("$\frac{1}{x^2}$", "$\frac{1}{x}$", "$\frac{1}{\sqrt{x}}$", "$\log{x}$", "$\sqrt{x}$", "$x$", "$x^2$"))

# knitr::kable(cbind(lambda=c(-2,-1,-0.5,0,0.5,1,2), Transformation=c("\$\\frac{1}{x^2}$", "\$\\frac{1}{x}$", "\$\\frac{1}{\\sqrt{x}}$", "\$\\log{x}$", "\$\\sqrt{x}$", "\$x$", "\$x^2$")), escape = F)
```

\begin{tabular}{l|r}
\hline
& lambda & Transformation\$\\
\hline
-2 & $\frac{1}{x^2}$
\hline
-1 & $\frac{1}{x}$
\hline
-0.5 & $\frac{1}{\sqrt{x}}$
\hline
0 & $\log{x}$
\hline
0.5 & $\sqrt{x}$
\hline
1 & $x$
\hline
2 & $x^2$
\end{tabular}

@@

#### Normalization
The term normalization is ambiguous in data-science and will be explained briefly in this chapter. Starting with converting the data to the Z-distribution (Gaussian result), where 0 reflects the mean and 1 refers to the standard deviation, until vectorization where every feature (=protein) can be explained as a vector which points in a specific direction in the unit sphere, normalization can have different meanings. Normalization is considered as part of the scaling procedure and ensures that every feature contributes equally to our statistical model and hinders large values from biasing the model in a particular direction. Although this could also be harmful for the modeling procedure, since the impact of important features on the dataset will also be reduced. 
Depending on the algorithms later used in the data processing, the method of choice for normalization can change the results significantly. The K-means clustering algorithm for instance relies on the distances between distinct data points by minimizing variance of the squared euclidean distances. This method will be applied in the missing value imputation which will be the next step applied to our data. Hence we want to make sure that the get the maximum accuracy for our imputed values. 
The first option to choose in the user interface is the column-wise median and row-wise mean normalization. This method was chosen by Specht et al in the SCoPE2 publication 2021 /citep{Specht2021} and worked as a reference during the development process to benchmark other methods. When dividing each value of an expression set by the median of the particular column, the procedure is considered as a column-wise median normalization. Hence the mean-normalization works the same way on the column. In the analysis pipeline both procedures were applied on the dataset. However when the logarithm as the transformation method was chosen in advance the pipeline automatically switches from a division to a subtraction, since:
$\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v$
